---
title: "Simplified data analysis"
author: "Dawson Shero"
date: "2026-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r}
library(here) 
library(readxl) 
library(janitor) 
library(tidyverse)
library(broom)
library(car)
library(lmtest)
library(sandwich)
library(emmeans)
library(interactions)
library(performance)
library(ggcorrplot)
library(rstatix)
library(psych)
library(modelsummary)
```

```{r cleaning, results='hide'}
df_raw <- read.csv(here("data", "combined_tests.csv"))

df_raw <- df_raw |> 
  rename(file = "source_file")

df_raw_sum <- read.csv(here("data", "Data_Summary_with_age.csv"))

df_raw_sum <- df_raw_sum |> 
  select("file", "age")

df_raw <- df_raw |> 
  group_by(file) |> 
  mutate(max_bla = max(bla))

### Making a df that only has variables that are needed for the study

colnames(df_raw)

df_clean <- df_raw |> 
  select("file", "gender", "event", "test", "height", "weight", "bmi", "fat", "lbm", "hb",
         "hct", "rbc", "wbc", "max_bla", "vo2max", "vo2max_absolute", "hrmax", "load_tlac",
         "hr_tlac", "percent_vo2max_tlac", "percent_h_rmax_tlac", "bla_tlac")

df_clean <- merge(df_clean, df_raw_sum, by = "file")

df_clean <- df_clean |> 
  relocate("age", .after = gender)

df_clean <- df_clean |> 
  distinct(file, .keep_all = T)

df_clean <- df_clean |>
  filter(!file %in% c("A539.xlsx", "A140.xlsx", "A562.xlsx", "A1065.xlsx"))

### Df manipulation and cleaning for analysis
colnames(df_clean)

anyNA(df_clean)

colSums(is.na(df_clean))

df_clean$weight[df_clean$weight == "Missing"] <- NA

df_clean <- df_clean |> 
  na.omit() |> 
  mutate(weight = as.numeric(as.character(weight)))

df_clean <- df_clean |> 
  mutate(across(where(is.numeric), ~ round(., 2)))

df_clean <- df_clean |> 
  filter(age >= 18)

df_clean <- df_clean |> 
  rename(sex = gender)

df_clean <- df_clean %>%
  mutate(test = case_match(test,
    c("Rowing", "Kayak") ~ "watersports",
    .default = test
  ))

df_clean <- df_clean |> 
  mutate(age_cat = case_when(
    age <= 30  ~ "under 30",
    age >= 30 ~ "over 30",
    TRUE ~ NA_character_
  )) |> 
  mutate(age_cat = factor(age_cat, 
                           levels = c("under 30", "over 30"),
                           ordered = TRUE)) |> 
  relocate(age_cat, .after = age)
```

### Is there a significant difference in absolute VO2max across GxT protocols

```{r}
## Setting test as factor to allow post-hoc
str(df_clean$test)
df_clean$test <- as.factor(df_clean$test)

# Statistical assumptions 

## Homogeneity of Variance
leveneTest(vo2max_absolute ~ test, data = df_clean)

##### Due the the p-value being lower than .05 in Levene test variance was not equal, thus changed the ANOVA into a Welch ANOVA for un-equal variance

## Normality (Shapiro-Wilk)
df_clean |> 
  group_by(test) |> 
  shapiro_test(vo2max_absolute)

##### Due to the large sample size we can still safely use Welch ANOVA due to the robustness to violations of mormality (Central Limit Therom)

## Look at skewness of test data
ggplot(df_clean, aes(x = vo2max_absolute, fill = test)) +
  geom_histogram(binwidth = 0.2, color = "black", alpha = 0.7) +
  facet_wrap(~ test, ncol = 1) +
  theme_minimal() +
  labs(
    title = "Absolute VO2max Distribution Across Test Modalities",
    x = "Absolute VO2max (L/min)",
    y = "Count"
  )

df_clean |> 
  group_by(test) |> 
  summarise(skewness = skew(vo2max_absolute),
    n = n()
  )

##### I ran this to make sure the there wasn't any skewness so that welch ANOVA could run properly

# Analysis (welch ANOVA + Post-Hoc)

## independent one way ANOVA (welch ANOVA)
aov_model <- oneway.test(vo2max_absolute ~ test, data = df_clean, var.equal = F)
aov_model

##### One-way test showed significance, now running post-hoc test to find which test is significant

## Games-Howell post-hoc
games_howell_test(data = df_clean, vo2max_absolute ~ test)

##### Games-Howell test showed statistically significant pairwise differences across all three test modalities
```

> This does show that there is statistically significant differences in the GxT protocols and absolute VO2max. This allows us to then go and look at how known predictors may change the associations of absolute VO2max based on the GxT protocol the researchers use. 

### how do known predictors for vo2max change based on the selection of exercise modalitiy

```{r}
# Building models of various predictors (block entry)

## Demographics
B1 <- lm(vo2max_absolute ~ sex + age, data = df_clean)
## + Body composition
B2 <- update(B1, . ~ . + height + fat + lbm)
## + Hematology
B3 <- update(B2, . ~ . + hb + hct + max_bla)
## + Test main effect
B4 <- update(B3, . ~ . + test)
## + Moderation
B5 <- update(B4, . ~ . + (height + fat + lbm + hb + hct + max_bla) * test) 
summary(B5)

## Does body composition improve the demographic model?
anova(B1, B2)

## Do hematological/metabolic variables improve the model?
anova(B2, B3)

## Does exercise modality improve the additive model?
anova(B3, B4)

## Overall test of whether predictor associations differ by modality
anova(B4, B5)

# Assumption checks

## The standard diagnostic plots
par(mfrow = c(2, 2))
plot(B5)
par(mfrow = c(1, 1))

## Multicollinearity
vif(B5)

## Homoscedasticity
bptest(B5)

# Individual interaction tests
drop1(B5, test = "F")

# What is the slope of the predictor within each protocol
vars <- c("height", "fat", "lbm", "hb", "hct", "max_bla")
purrr::map(purrr::set_names(vars), ~ emtrends(B5, ~ test, var = .x, infer = TRUE))

##### I decided that a step regression is not the right regression since that uses automated statistical algorithms to select variables. I went with a Hierarchical regression (block entry) since we know the known predictors.
```


### Visualizations

```{r}
r2_tbl <- map_dfr(
  list("1: Demographics" = B1, "2: + Body comp" = B2, "3: + Hematology" = B3,
       "4: + Test" = B4, "5: + Moderation" = B5),
  ~ glance(.x) |> dplyr::select(r.squared, adj.r.squared),
  .id = "model"
) |>
  mutate(model = fct_inorder(model))

ggplot(r2_tbl, aes(x = model, y = adj.r.squared)) +
  geom_col(fill = "#0072B2", width = 0.7) +
  geom_text(aes(label = scales::percent(adj.r.squared, accuracy = 0.1)),
            vjust = -0.4, size = 3.5) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal(base_size = 12) +
  labs(title = "Variance in absolute VO2max explained by each model block",
       x = NULL, y = "Adjusted R²")
```

```{r}
ggplot(df_clean, aes(x = test, fill = test)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Participants by Exercise Modality",
    x = "Exercise Modality",
    y = "Number of Participants"
  )
```

```{r}
ggplot(df_clean, aes(x = test, fill = sex)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Sex Distribution Across Exercise Modalities",
    x = "Exercise Modality",
    y = "Number of Participants",
    fill = "Sex"
  )
```

```{r}
ggplot(df_clean, aes(x = test, y = age)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Age Distribution Across Exercise Modalities",
    x = "Exercise Modality",
    y = "Age (years)"
  )
```

```{r}
ggplot(df_clean, aes(x = test, y = vo2max_absolute)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Absolute VO₂max Across Exercise Modalities",
    x = "Exercise Modality",
    y = "Absolute VO₂max (L/min)"
  )
```




